---
title: "Stat 479 Group Project with Matrix and Multiple Logistic Regression"
author: "Basketball Playoff Prediction Team"
date: "Nov 22 2021"
output:
  html_document:
    df_print: paged
header-includes:
- \usepackage{fancyhdr}
- \pagestyle{fancy}
- \fancyhf{}
- \rhead{BPP}
- \lhead{STAT 479 GP}
- \cfoot{\thepage}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, comment=NA, tidy=TRUE, tidy.opts=list(width.cutoff=45), warning=FALSE, message=FALSE, fig.align='center')

library(tidyverse)
library(dplyr)
library(plyr)
library(knitr)
library(rstan)
```

# Merging Data and Data Cleaning

```{r import data}
Regular <- read.csv("cleaned_data_v1_12_18.csv") # import the data 
Regular$Success <- as.integer(Regular$Y) # add a column of True = 1 and False = 0 from Column "Y"
y <- Regular$Success
deltaOEFF <- Regular[, "deltaOEFF"]
deltaDEFF <- Regular[, "deltaDEFF"]
summary(deltaOEFF) # let the grid for be [-9, 9]
summary(deltaDEFF) # let the grid for be [-9, 9]
```

## Model

```{r model fitting}
y <- Regular$Success # length = 502
deltaOEFF_grid <- seq(-9, 9, by = 0.1) # grid for posterior prediction
deltaDEFF_grid <- seq(-9, 9, by = 0.1) # length = 181
x <- cbind(deltaOEFF, deltaDEFF) # dim(584, 2)
K <- 1
N <- length(y) # N = 502
D <- dim(x)[2] # D = 2
n_grid = length(deltaOEFF_grid) # n_grid = 181
data_list <- list(K = K, # create a list to fill rstan model
                  N = N,
                  D = D,
                  y = y,
                  x = x, 
                  n_grid = n_grid, 
                  deltaOEFF_grid = deltaOEFF_grid, 
                  deltaDEFF_grid = deltaDEFF_grid)
test_stan <- stan_model(file = "multi_logistic.stan")
fit <- sampling(object = test_stan, data = data_list)
summary(fit)[[1]][,"Rhat"]
```

# Preliminary visual plots

```{r posterior coef}
beta1_samples <- rstan::extract(fit, pars = "beta[1,1]")[["beta[1,1]"]] # extract 4000 coefficients 
beta2_samples <- rstan::extract(fit, pars = "beta[2,1]")[["beta[2,1]"]]
mean(beta1_samples)
mean(beta2_samples)
```

As we tune the prior of beta_1 and beta_2, the posterior betas do not change very much. As long as the prior stays within a reasonable range like keeping beta_1 positive and beta_2 negative, the posterior coefficients will stay within the range where beta_1 is around 0.25 and beta_2 around -0.15. This is coherent with the Bayesian characteristic that the posterior probability will be close to the likelihood / observed data if given sufficient data. In our case, we have 584 observed data, which can be considered significantly large.

## Plots for coefficients:

```{r coef plots, fig.height = 7, fig.width = 12, fig.align = "center"}
par(mfrow = c(1,2))
hist(beta1_samples, main = "beta_1", breaks = 100, xlab = "beta_1")
hist(beta2_samples, main = "beta_2", breaks = 100, xlab = "beta_2")
```

# Posterior Model Plots

Since we have 2 predictors, the ideal plot shall be 3D. In order to present the plot in 2D, we fix one predictor on certain values and put the other predictor on the x-axis. In order to present the data properly, we decide to fix one predictor on 3 values: its mean, mean + 1sd, mean - 1sd.

## Plot the predictive probabilities

```{r fig.height = 8, fig.width = 12, fig.align = "center"}
post_grid_meanD <- rstan::extract(fit, pars = "prob_meanD")[["prob_meanD"]]
prob_meanD_minus_sd <- rstan::extract(fit, pars = "prob_meanD_minus_sd")[["prob_meanD_minus_sd"]]
prob_meanD_plus_sd <- rstan::extract(fit, pars = "prob_meanD_plus_sd")[["prob_meanD_plus_sd"]]
# at mean deltaDEFF
post_grid_meanD_mean <- apply(post_grid_meanD, MARGIN = 2, FUN = mean)
post_grid_meanD_l95 <- apply(post_grid_meanD, MARGIN = 2, FUN = quantile, probs = 0.025)
post_grid_meanD_u95 <- apply(post_grid_meanD, MARGIN = 2, FUN = quantile, probs = 0.975)
# at mean deltaDEFF - sd
prob_meanD_minus_sd_mean <- apply(prob_meanD_minus_sd, MARGIN = 2, FUN = mean)
prob_meanD_minus_sd_l95 <- apply(prob_meanD_minus_sd, MARGIN = 2, FUN = quantile, probs = 0.025)
prob_meanD_minus_sd_u95 <- apply(prob_meanD_minus_sd, MARGIN = 2, FUN = quantile, probs = 0.975)
# at mean deltaDEFF + sd
prob_meanD_plus_sd_mean <- apply(prob_meanD_plus_sd, MARGIN = 2, FUN = mean)
prob_meanD_plus_sd_l95 <- apply(prob_meanD_plus_sd, MARGIN = 2, FUN = quantile, probs = 0.025)
prob_meanD_plus_sd_u95 <- apply(prob_meanD_plus_sd, MARGIN = 2, FUN = quantile, probs = 0.975)
# plot
plot(1, type= "n", xlim = c(-9, 9), ylim = c(0,1),
     main = "Probability with OEFF and fixed DEFF", xlab = "deltaOEFF", ylab = "Probability")
# mean
polygon(x = c(deltaOEFF_grid, rev(deltaOEFF_grid)), 
        y = c(post_grid_meanD_l95, rev(post_grid_meanD_u95)),
        col = alpha("black", 0.8),
        border = NA)
lines(deltaOEFF_grid, post_grid_meanD_mean, lwd = 1.5)
# mean - sd 
polygon(x = c(deltaOEFF_grid, rev(deltaOEFF_grid)), 
        y = c(prob_meanD_minus_sd_l95, rev(prob_meanD_minus_sd_u95)),
        col = alpha("grey", 0.7),
        border = NA)
lines(deltaOEFF_grid, prob_meanD_minus_sd_mean, lwd = 1.5)
# mean + sd
polygon(x = c(deltaOEFF_grid, rev(deltaOEFF_grid)), 
        y = c(prob_meanD_plus_sd_l95, rev(prob_meanD_plus_sd_u95)),
        col = alpha("red", 0.6),
        border = NA)
lines(deltaOEFF_grid, prob_meanD_plus_sd_mean, lwd = 1.5)
legend("bottomright", legend = c("mean_deltaDEFF", "mean_minus_1sd", "mean_plus_1sd"), 
       lty = c(1, 1, 1), lwd = c(2, 2, 2), col = c('black', 'grey', 'red'))
```

The first plot uses all 4000 beta_1s and beta_2s as coefficients of the logistic regression. Here we can see that the x-axis deltaOEFF is positively associated with the posterior probability, where the black line is predictor deltaOEFF when deltaDEFF is at its mean, the grey line is deltaDEFF at mean minus one standard deviation, and the red line is deltaDEFF at mean plus one standard deviation. We can see the grey line generally has a higher probability (the position is mean - 1sd). This may be because the two predictors tend to be oppositely plotted, which makes sense in real life as the offensive rate is calculated as the opposite of the other team's defensive rate.

Now try to fix deltaOEFF and put deltaDEFF on x-axis

```{r fig.height = 8, fig.width = 12, fig.align = "center"}
post_grid_meanO <- rstan::extract(fit, pars = "prob_meanO")[["prob_meanO"]]
prob_meanO_minus_sd <- rstan::extract(fit, pars = "prob_meanO_minus_sd")[["prob_meanO_minus_sd"]]
prob_meanO_plus_sd <- rstan::extract(fit, pars = "prob_meanO_plus_sd")[["prob_meanO_plus_sd"]]
# at mean deltaOEFF
post_grid_meanO_mean <- apply(post_grid_meanO, MARGIN = 2, FUN = mean)
post_grid_meanO_l95 <- apply(post_grid_meanO, MARGIN = 2, FUN = quantile, probs = 0.025)
post_grid_meanO_u95 <- apply(post_grid_meanO, MARGIN = 2, FUN = quantile, probs = 0.975)
# at mean deltaOEFF - sd
prob_meanO_minus_sd_mean <- apply(prob_meanO_minus_sd, MARGIN = 2, FUN = mean)
prob_meanO_minus_sd_l95 <- apply(prob_meanO_minus_sd, MARGIN = 2, FUN = quantile, probs = 0.025)
prob_meanO_minus_sd_u95 <- apply(prob_meanO_minus_sd, MARGIN = 2, FUN = quantile, probs = 0.975)
# at mean deltaOEFF + sd
prob_meanO_plus_sd_mean <- apply(prob_meanO_plus_sd, MARGIN = 2, FUN = mean)
prob_meanO_plus_sd_l95 <- apply(prob_meanO_plus_sd, MARGIN = 2, FUN = quantile, probs = 0.025)
prob_meanO_plus_sd_u95 <- apply(prob_meanO_plus_sd, MARGIN = 2, FUN = quantile, probs = 0.975)
# plot
plot(1, type= "n", xlim = c(-9, 9), ylim = c(0,1),
     main = "Probability with DEFF and fixed OEFF", xlab = "deltaDEFF", ylab = "Probability")
# mean
polygon(x = c(deltaDEFF_grid, rev(deltaDEFF_grid)), 
        y = c(post_grid_meanO_l95, rev(post_grid_meanO_u95)),
        col = alpha("black", 0.8),
        border = NA)
lines(deltaDEFF_grid, post_grid_meanO_mean, lwd = 1.5)
# mean - sd 
polygon(x = c(deltaDEFF_grid, rev(deltaDEFF_grid)), 
        y = c(prob_meanO_minus_sd_l95, rev(prob_meanO_minus_sd_u95)),
        col = alpha("grey", 0.7),
        border = NA)
lines(deltaDEFF_grid, prob_meanO_minus_sd_mean, lwd = 1.5)
# mean + sd
polygon(x = c(deltaDEFF_grid, rev(deltaDEFF_grid)), 
        y = c(prob_meanO_plus_sd_l95, rev(prob_meanO_plus_sd_u95)),
        col = alpha("red", 0.6),
        border = NA)
lines(deltaDEFF_grid, prob_meanO_plus_sd_mean, lwd = 1.5)
legend("bottomleft", legend = c("mean_deltaOEFF", "mean_minus_1sd", "mean_plus_1sd"), 
       lty = c(1, 1, 1), lwd = c(2, 2, 2), col = c('black', 'grey', 'red'))
```

The probability of winning is roughly negatively associated with the predictor deltaDEFF.

# Zsun: Simulation

```{r}
source("zsun_playoff_simulator.R")

set.seed(20211129)
pairwise_data <- read.csv("pairwise_DEFF_OEFF_2019.csv")
champions_4000 <- rep()
prob_matrix_4000 <- list()
for (i in 1:4000){
  simu_i_prob <- pairwise_data %>% 
    mutate(p = exp(d_OEFF * beta1_samples[i] + d_DEFF * beta2_samples[i])/ (1 + exp(d_OEFF * beta1_samples[i] + d_DEFF * beta2_samples[i])))
  prob_matrix_i <- matrix(simu_i_prob$p,nrow=16,ncol=16)
  name <- pairwise_data[1:16,1]
  colnames(prob_matrix_i) <- name
  rownames(prob_matrix_i) <- name
  prob_matrix_4000[[i]] <- prob_matrix_i # store the probability matrix
  champions_4000[i] <- simu_final(P = prob_matrix_i)[1] # store the simu_result
}
table(champions_4000)/4000
# length(prob_matrix_4000)
```

# MSE

## MSE for 18-19 playoff prediction

```{r}
Regular_19 <- read.csv("cleaned_data_v1_19.csv") # import the data 
Regular_19$Success <- as.integer(Regular_19$Y)
deltaOEFF_19 <- Regular_19$deltaOEFF
deltaDEFF_19 <- Regular_19$deltaDEFF
predicted <- mean(beta1_samples) * deltaOEFF_19 + mean(beta2_samples) * deltaDEFF_19
pred_prob <- exp(predicted)/(exp(predicted) + 1)
observed <- Regular_19$Success
O_minus_P <- (observed - pred_prob)^2
sum(O_minus_P)/length(pred_prob)
```

# Cross validation

## Split into 5 folds

```{r}
set.seed(123)
idx <- sample(1:502)
fold1 <- idx[1:100]
fold2 <- idx[101:200]
fold3 <- idx[201:300]
fold4 <- idx[301:401]
fold5 <- idx[402:502]
```

## MSE using DEFF only

### fold 1

```{r}
train_cv1 <- Regular[-fold1,]
test_cv1 <- Regular[fold1,]

y <- train_cv1$Success # length = 584
deltaDEFF_grid <- seq(-9, 9, by = 0.1) # grid for prediction
x <- train_cv1$deltaDEFF
N <- length(y) 
n_grid = length(deltaDEFF_grid) # n_grid = 181
data_list <- list(n_grid = n_grid,
                  y = y, 
                  deltaDEFF = x, 
                  N = N, 
                  deltaDEFF_grid = deltaDEFF_grid)
test_stan <- stan_model(file = "DEFF_Only.stan")
fit <- sampling(object = test_stan, data = data_list)
beta2_samples <- extract(fit, pars = "beta")[["beta"]] # extract 4000 coefficients each
### Calculate MSE
mean.p.train <- rep(0,dim(train_cv1)[1])
mean.p.test <- rep(0,dim(test_cv1)[1])
for ( i in 1:4000){
  b2 <- beta2_samples[i]
  log.pred.train <- b2 * train_cv1$deltaDEFF
  log.pred.test <- b2 * test_cv1$deltaDEFF
  p.train <- exp(log.pred.train) / (1 + exp(log.pred.train))
  p.test <- exp(log.pred.test) / (1 + exp(log.pred.test))
  mean.p.train <- mean.p.train + p.train/4000
  mean.p.test <- mean.p.test  + p.test/4000
}

(mse_cv1_train <- mean((train_cv1$Y - mean.p.train)^2))
(mse_cv1_test <- mean((test_cv1$Y - mean.p.test)^2))
```

### fold 2

```{r}
train_cv2 <- Regular[-fold2,]
test_cv2 <- Regular[fold2,]

y <- train_cv2$Success 
deltaDEFF_grid <- seq(-9, 9, by = 0.1) 
x <- train_cv2$deltaDEFF
N <- length(y) 
n_grid = length(deltaDEFF_grid) 
data_list <- list(n_grid = n_grid,
                  y = y, 
                  deltaDEFF = x, 
                  N = N, 
                  deltaDEFF_grid = deltaDEFF_grid)
test_stan <- stan_model(file = "DEFF_Only.stan")
fit <- sampling(object = test_stan, data = data_list)
beta2_samples <- extract(fit, pars = "beta")[["beta"]] 
### Calculate MSE
mean.p.train <- rep(0,dim(train_cv2)[1])
mean.p.test <- rep(0,dim(test_cv2)[1])
for (i in 1:4000){
  b2 <- beta2_samples[i]
  log.pred.train <- b2 * train_cv2$deltaDEFF
  log.pred.test <- b2 * test_cv2$deltaDEFF
  p.train <- exp(log.pred.train) / (1 + exp(log.pred.train))
  p.test <- exp(log.pred.test) / (1 + exp(log.pred.test))
  mean.p.train <- mean.p.train + p.train/4000
  mean.p.test <- mean.p.test  + p.test/4000
}

(mse_cv2_train <- mean((train_cv2$Y - mean.p.train)^2))
(mse_cv2_test <- mean((test_cv2$Y - mean.p.test)^2))
```

### fold 3

```{r}
train_cv3 <- Regular[-fold3,]
test_cv3 <- Regular[fold3,]

y <- train_cv3$Success 
deltaDEFF_grid <- seq(-9, 9, by = 0.1) 
x <- train_cv3$deltaDEFF
N <- length(y) 
n_grid = length(deltaDEFF_grid) 
data_list <- list(n_grid = n_grid,
                  y = y, 
                  deltaDEFF = x, 
                  N = N, 
                  deltaDEFF_grid = deltaDEFF_grid)
test_stan <- stan_model(file = "DEFF_Only.stan")
fit <- sampling(object = test_stan, data = data_list)
beta2_samples <- extract(fit, pars = "beta")[["beta"]] 
### Calculate MSE
mean.p.train <- rep(0,dim(train_cv3)[1])
mean.p.test <- rep(0,dim(test_cv3)[1])
for (i in 1:4000){
  b2 <- beta2_samples[i]
  log.pred.train <- b2 * train_cv3$deltaDEFF
  log.pred.test <- b2 * test_cv3$deltaDEFF
  p.train <- exp(log.pred.train) / (1 + exp(log.pred.train))
  p.test <- exp(log.pred.test) / (1 + exp(log.pred.test))
  mean.p.train <- mean.p.train + p.train/4000
  mean.p.test <- mean.p.test  + p.test/4000
}

(mse_cv3_train <- mean((train_cv3$Y - mean.p.train)^2))
(mse_cv3_test <- mean((test_cv3$Y - mean.p.test)^2))
```

### fold 4

```{r}
train_cv4 <- Regular[-fold4,]
test_cv4 <- Regular[fold4,]

y <- train_cv4$Success 
deltaDEFF_grid <- seq(-9, 9, by = 0.1) 
x <- train_cv4$deltaDEFF
N <- length(y) 
n_grid = length(deltaDEFF_grid) 
data_list <- list(n_grid = n_grid,
                  y = y, 
                  deltaDEFF = x, 
                  N = N, 
                  deltaDEFF_grid = deltaDEFF_grid)
test_stan <- stan_model(file = "DEFF_Only.stan")
fit <- sampling(object = test_stan, data = data_list)
beta2_samples <- extract(fit, pars = "beta")[["beta"]] 
### Calculate MSE
mean.p.train <- rep(0,dim(train_cv4)[1])
mean.p.test <- rep(0,dim(test_cv4)[1])
for (i in 1:4000){
  b2 <- beta2_samples[i]
  log.pred.train <- b2 * train_cv4$deltaDEFF
  log.pred.test <- b2 * test_cv4$deltaDEFF
  p.train <- exp(log.pred.train) / (1 + exp(log.pred.train))
  p.test <- exp(log.pred.test) / (1 + exp(log.pred.test))
  mean.p.train <- mean.p.train + p.train/4000
  mean.p.test <- mean.p.test  + p.test/4000
}

(mse_cv4_train <- mean((train_cv4$Y - mean.p.train)^2))
(mse_cv4_test <- mean((test_cv4$Y - mean.p.test)^2))
```

### fold 5 

```{r}
train_cv5 <- Regular[-fold5,]
test_cv5 <- Regular[fold5,]

y <- train_cv4$Success 
deltaDEFF_grid <- seq(-9, 9, by = 0.1) 
x <- train_cv5$deltaDEFF
N <- length(y) 
n_grid = length(deltaDEFF_grid) 
data_list <- list(n_grid = n_grid,
                  y = y, 
                  deltaDEFF = x, 
                  N = N, 
                  deltaDEFF_grid = deltaDEFF_grid)
test_stan <- stan_model(file = "DEFF_Only.stan")
fit <- sampling(object = test_stan, data = data_list)
beta2_samples <- extract(fit, pars = "beta")[["beta"]] 
### Calculate MSE
mean.p.train <- rep(0,dim(train_cv5)[1])
mean.p.test <- rep(0,dim(test_cv5)[1])
for (i in 1:4000){
  b2 <- beta2_samples[i]
  log.pred.train <- b2 * train_cv5$deltaDEFF
  log.pred.test <- b2 * test_cv5$deltaDEFF
  p.train <- exp(log.pred.train) / (1 + exp(log.pred.train))
  p.test <- exp(log.pred.test) / (1 + exp(log.pred.test))
  mean.p.train <- mean.p.train + p.train/4000
  mean.p.test <- mean.p.test  + p.test/4000
}

(mse_cv5_train <- mean((train_cv5$Y - mean.p.train)^2))
(mse_cv5_test <- mean((test_cv5$Y - mean.p.test)^2))
```

```{r}
# train
mean(c(mse_cv1_train,mse_cv2_train,mse_cv3_train,mse_cv4_train,mse_cv5_train))
# test
mean(c(mse_cv1_test,mse_cv2_test,mse_cv3_test,mse_cv4_test,mse_cv5_test))
```
